---
title: "Extreme Value Theory"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r warning=FALSE,message=FALSE}
library(rnoaa)
library(leaflet)
library(lawn)
options(noaakey = "piRauSeecCMaIVkleStduqdxoaLLOPMs")
lawn_bbox_polygon(c(113.338953078, -43.6345972634, 153.569469029, -10.6681857235)) %>% view
# List stations corresponding to Australia
all_stations<-ncdc_stations(extent = c(-43.6345972634, 113.338953078, -10.6681857235, 153.569469029))
head(all_stations)
```

# Get Australian Rainfall data

wrapper_get_meta_data function get all Australian station data from R package "rnoaa".

```{r warning=FALSE,message=FALSE}
wrapper_get_meta_data <- function(country_id = "AS", network_id = "N"){

  ghcnd_meta_data = ghcnd_stations()
  as_meta_data = dplyr::filter(ghcnd_meta_data,
                               substr(id, 1, 2) == country_id & substr(id, 3, 3) == network_id)

  return(as_meta_data)

}
```


```{r warning=FALSE,message=FALSE}
as_meta_data = wrapper_get_meta_data()
save(as_meta_data, file = "Data/AS_meta_data.RData")
head(as_meta_data)

```
Above meta_data contains following precipitation_elements.\\

* PRCP is the daily precipitation (**tenths of a mm**)  
* MDPR is a multiday accumulated total (tenths of a mm)  
* DAPR is the number of days in a multiday total  
* DWPR number of days of non-zero rainfall in a multiday total. 

Hence, we filter the PRCP only for further analysis.

```{r warning=FALSE,message=FALSE}
# Filtering pricipitation data
prcp_meta_data = dplyr::filter(as_meta_data, element == "PRCP")
head(prcp_meta_data)

```

From this data set, distribution of stations are visualized through map visualization to get an idea of spreading of rain gauges in Australia.

```{r warning=FALSE,message=FALSE}
library(ggplot2)
ggplot(prcp_meta_data, aes(x = longitude, y = latitude)) + geom_point(size = 0.2,col="blue")
```

I have selected eight sites to reflect the rainfall categories as follows:\\

* Birdsville Police Station (38002) [Arid] 
* Cambooya Post Office (41011) [Summer] 
* Cape Naturaliste (9519) [Winter Dominant] 
* Cataract Dam (68016) [Uniform] 
* Eudunda (24511) [Winter] 
* Gabo Island Lighthouse (84016) [Uniform] 
* Oenpelli (14042) [Summer Dominant] 
* Sorell (Whitlea) (94063) [Uniform]

```{r warning=FALSE,message=FALSE}
# function for obtaining precipitation data

Australian_prcp_data <- function(stn_ids){
  prcp_data <-meteo_pull_monitors(monitors=stn_ids)
  prcp_data <- select(prcp_data,c('id','date','prcp'))
  return(prcp_data)
}

```

```{r warning=FALSE,message=FALSE}
library(dplyr)
station_id="ASN00038002"
# Obtain the corresponding precipitation data
prcp_data_38002 <- Australian_prcp_data(station_id)
prcp_data_38002<-prcp_data_38002 %>% filter(date<="2022-06-04")
#head(prcp_data_38002)
station_id="ASN00041011"
# Obtain the corresponding precipitation data
prcp_data_41011 <- Australian_prcp_data(station_id)
#head(prcp_data_41011)
station_id="ASN00009519"
# Obtain the corresponding precipitation data
prcp_data_9519 <- Australian_prcp_data(station_id)
prcp_data_9519<-prcp_data_9519 %>% filter(date<="2022-07-07")
#head(prcp_data_9519)
station_id="ASN00068016"
# Obtain the corresponding precipitation data
prcp_data_68016 <- Australian_prcp_data(station_id)
#head(prcp_data_68016)
station_id="ASN00024511"
# Obtain the corresponding precipitation data
prcp_data_24511 <- Australian_prcp_data(station_id)
prcp_data_24511<-prcp_data_24511 %>% filter(date<="2022-07-07")
#head(prcp_data_24511)
station_id="ASN00084016"
# Obtain the corresponding precipitation data
prcp_data_84016 <- Australian_prcp_data(station_id)
prcp_data_84016<-prcp_data_84016 %>% filter(date<="2022-07-07")
#head(prcp_data_84016)
station_id="ASN00014042"
# Obtain the corresponding precipitation data
prcp_data_14042 <- Australian_prcp_data(station_id)
#head(prcp_data_14042)
station_id="ASN00094063"
# Obtain the corresponding precipitation data
prcp_data_94063 <- Australian_prcp_data(station_id)
#head(prcp_data_94063)

```
### Visualizing rainfall data

```{r warning=FALSE,message=FALSE}
a<-ggplot(prcp_data_38002, aes(x=date, y=prcp))+ geom_point(col="#1b9e77")+ggtitle("Gauge number:38002")+ theme(plot.title=element_text(hjust=0.5))

b<-ggplot(prcp_data_41011, aes(x=date, y=prcp))+ geom_point(col="#1b9e77")+ggtitle("Gauge number:41011")+ theme(plot.title=element_text(hjust=0.5))

c<-ggplot(prcp_data_9519, aes(x=date, y=prcp))+ geom_point(col="#1b9e77")+ggtitle("Gauge number:9519")+ theme(plot.title=element_text(hjust=0.5))

d<-ggplot(prcp_data_68016, aes(x=date, y=prcp))+ geom_point(col="#1b9e77")+ggtitle("Gauge number:68016")+ theme(plot.title=element_text(hjust=0.5))

e<-ggplot(prcp_data_24511, aes(x=date, y=prcp))+ geom_point(col="#1b9e77")+ggtitle("Gauge number:24511")+ theme(plot.title=element_text(hjust=0.5))

f<-ggplot(prcp_data_84016, aes(x=date, y=prcp))+ geom_point(col="#1b9e77")+ggtitle("Gauge number:84016")+ theme(plot.title=element_text(hjust=0.5))

g<-ggplot(prcp_data_14042, aes(x=date, y=prcp))+ geom_point(col="#1b9e77")+ggtitle("Gauge number:14042")+ theme(plot.title=element_text(hjust=0.5))

h<-ggplot(prcp_data_94063, aes(x=date, y=prcp))+ geom_point(col="#1b9e77")+ggtitle("Gauge number:94063")+ theme(plot.title=element_text(hjust=0.5))

library(patchwork)

plts1<-(a|b)/(c|d)
plts1
plts2<-(e|f)/(g|h)
plts2
```

# **Block Maxima Approach - Generalized Extreme Value Distribution (GEV)**

```{r warning=FALSE,message=FALSE}
Annual_maximum_rainfall<-function(data){
 yr1= min(lubridate::year(data$date))
 yr2= max(lubridate::year(data$date))
 n=yr2-yr1+1
 annmax_prcp <- matrix(data = NA,nrow=n,ncol = 1)
 for (i in 1:n){
   yr = (yr1-1)+i
   index = which((lubridate::year(data$date)==yr))
   yr_wise_rainfall<-data[index,3]
   annmax_prcp[i,1] <- max(yr_wise_rainfall,na.rm = TRUE)
   #annmax_prcp[i,2] <- yr
 }
 return(annmax_prcp)
}

```


```{r warning=FALSE,message=FALSE}
Max_prcp_data_38002<-Annual_maximum_rainfall(prcp_data_38002)
Max_prcp_data_41011<-Annual_maximum_rainfall(prcp_data_41011)
Max_prcp_data_9519<-Annual_maximum_rainfall(prcp_data_9519)
Max_prcp_data_68016<-Annual_maximum_rainfall(prcp_data_68016)
Max_prcp_data_24511<-Annual_maximum_rainfall(prcp_data_24511)
Max_prcp_data_84016<-Annual_maximum_rainfall(prcp_data_84016)
Max_prcp_data_14042<-Annual_maximum_rainfall(prcp_data_14042)
Max_prcp_data_94063<-Annual_maximum_rainfall(prcp_data_94063)

```

#### Augmented Dickey-Fuller Test 

* check whether maximum rainfall value series are stationary or not.

H0: The time series is regarded as non-stationary.
H1: The time series is regarded as stationary.

```{r warning=FALSE,message=FALSE}
library(tseries)
adf.test(Max_prcp_data_38002[!is.infinite(rowSums(Max_prcp_data_38002)),])
# 38002 max rainfall series is stationary by ADF test
adf.test(Max_prcp_data_41011)
# 41011 max rainfall series is not stationary
adf.test(Max_prcp_data_9519[!is.infinite(rowSums(Max_prcp_data_9519)),])
# 9519 max rainfall series is stationary by ADF test
adf.test(Max_prcp_data_68016[!is.infinite(rowSums(Max_prcp_data_68016)),])
# 68016 max rainfall series is not stationary by ADF test
adf.test(Max_prcp_data_24511[!is.infinite(rowSums(Max_prcp_data_24511)),])
# 24511 max rainfall series is stationary by ADF test
adf.test(Max_prcp_data_84016[!is.infinite(rowSums(Max_prcp_data_84016)),])
# 84016 max rainfall series is stationary by ADF test
adf.test(Max_prcp_data_14042)
# 14042 max rainfall series is stationary by ADF test
adf.test(Max_prcp_data_94063[!is.infinite(rowSums(Max_prcp_data_94063)),])
# 94063 max rainfall series is stationary by ADF test

```

#### Plotting maximum value rainfall

```{r echo=FALSE, warning=FALSE,message=FALSE}
# Obtaining yearly series required for plotting
Yearly_Time<-function(data){
 yr1= min(lubridate::year(data$date))
 yr2= max(lubridate::year(data$date))
 year = c(seq(yr1,yr2,1))
 return(year)
}
```


```{r echo=FALSE, warning=FALSE,message=FALSE}
Year_data_38002<-Yearly_Time(prcp_data_38002)
Year_data_41011<-Yearly_Time(prcp_data_41011)
Year_data_9519<-Yearly_Time(prcp_data_9519)
Year_data_68016<-Yearly_Time(prcp_data_68016)
Year_data_24511<-Yearly_Time(prcp_data_24511)
Year_data_84016<-Yearly_Time(prcp_data_84016)
Year_data_14042<-Yearly_Time(prcp_data_14042)
Year_data_94063<-Yearly_Time(prcp_data_94063)
```

```{r echo=FALSE, warning=FALSE,message=FALSE}
df_max_rainfall__38002<-data.frame(Max_prcp_data_38002,Year_data_38002)
df_max_rainfall__41011<-data.frame(Max_prcp_data_41011,Year_data_41011)
df_max_rainfall__9519<-data.frame(Max_prcp_data_9519,Year_data_9519)
df_max_rainfall__68016<-data.frame(Max_prcp_data_68016,Year_data_68016)
df_max_rainfall__24511<-data.frame(Max_prcp_data_24511,Year_data_24511)
df_max_rainfall__84016<-data.frame(Max_prcp_data_84016,Year_data_84016)
df_max_rainfall__14042<-data.frame(Max_prcp_data_14042,Year_data_14042)
df_max_rainfall__94063<-data.frame(Max_prcp_data_94063,Year_data_94063)
```

* Best fit lines are indicated by black lines.

```{r echo=FALSE,warning=FALSE,message=FALSE}
a1<-ggplot(df_max_rainfall__38002, aes(x=Year_data_38002, y=Max_prcp_data_38002))+ geom_line(col="#1b9e77")+ggtitle("Gauge number:38002")+ theme(plot.title=element_text(hjust=0.5))+
geom_smooth(method=lm, se=FALSE,col="black",size=0.5)

b1<-ggplot(df_max_rainfall__41011, aes(x=Year_data_41011, y=Max_prcp_data_41011))+ geom_line(col="#1b9e77")+ggtitle("Gauge number:41011")+ theme(plot.title=element_text(hjust=0.5))+
geom_smooth(method=lm, se=FALSE,col="black",size=0.5)

c1<-ggplot(df_max_rainfall__9519, aes(x=Year_data_9519, y=Max_prcp_data_9519))+ geom_line(col="#1b9e77")+ggtitle("Gauge number:9519")+ theme(plot.title=element_text(hjust=0.5))+
geom_smooth(method=lm, se=FALSE,col="black",size=0.5)

d1<-ggplot(df_max_rainfall__68016, aes(x=Year_data_68016, y=Max_prcp_data_68016))+ geom_line(col="#1b9e77")+ggtitle("Gauge number:68016")+ theme(plot.title=element_text(hjust=0.5))+
geom_smooth(method=lm, se=FALSE,col="black",size=0.5)

plts_max_1<-(a1|b1)/(c1|d1)
plts_max_1

```

```{r echo=FALSE,warning=FALSE,message=FALSE}
e1<-ggplot(df_max_rainfall__24511, aes(x=Year_data_24511, y=Max_prcp_data_24511))+ geom_line(col="#1b9e77")+ggtitle("Gauge number:24511")+ theme(plot.title=element_text(hjust=0.5))+
geom_smooth(method=lm, se=FALSE,col="black",size=0.5)

f1<-ggplot(df_max_rainfall__84016, aes(x=Year_data_84016, y=Max_prcp_data_84016))+ geom_line(col="#1b9e77")+ggtitle("Gauge number:84016")+ theme(plot.title=element_text(hjust=0.5))+
geom_smooth(method=lm, se=FALSE,col="black",size=0.5)

g1<-ggplot(df_max_rainfall__14042, aes(x=Year_data_14042, y=Max_prcp_data_14042))+ geom_line(col="#1b9e77")+ggtitle("Gauge number:14042")+ theme(plot.title=element_text(hjust=0.5))+
geom_smooth(method=lm, se=FALSE,col="black",size=0.5)

h1<-ggplot(df_max_rainfall__94063, aes(x=Year_data_94063, y=Max_prcp_data_94063))+ geom_line(col="#1b9e77")+ggtitle("Gauge number:94063")+ theme(plot.title=element_text(hjust=0.5))+
geom_smooth(method=lm, se=FALSE,col="black",size=0.5)

plts_max_2<-(e1|f1)/(g1|h1)
plts_max_2

```

##### Block Maxima is usually modelled using the distribution from Generalized Extreme Value Families such as Gumbell, Weibull, and Frechet.
* CDFs are as follows:

$$F(x)=exp[\ -(1-\xi \dfrac{x-\mu}{\sigma}) ]\; \hspace{0.5in} \xi \neq 0 $$

$$F(x)=exp[\ -exp(- \dfrac{x-\mu}{\sigma})]\ ; \hspace{0.5in} \xi =0$$

* location parameter - $\mu ; \hspace{0.1in} - \infty < \mu < \infty$

* scale parameter - $\sigma ; \hspace{0.1in} 0 < \sigma < \infty$ 

* shape parameter - $\xi \hspace{0.1in} - \infty < \xi < \infty$ 

* Gumbell distribution - $\xi=0$, Frechet distribution- $\xi >0$ , Weibull distribution - $\xi < 0$

$$Return \hspace{0.1in}Period=\dfrac{1}{P(Exceedance)}=\dfrac{1}{P(Z>z)}=\dfrac{1}{1-F(z)}$$
The significance of the best non-stationary model, when compared to the stationary model, can be measured from the p-value of the Chi-square distribution.

* negative log-likelihood of stationary model -nlls

* best non-stationary model -nllns

* 2[nlls − nllns] ∼ χ2(df)

##### Case 1
$$X_{t} \sim GEV(\mu, \sigma, \xi)$$
```{r warning=FALSE,message=FALSE}
library(ismev)
df<-Max_prcp_data_38002
df <- df[!is.infinite(rowSums(df)),]
A = gev.fit(df,ydat = NULL)
gev.diag(A)
```

##### Case 2
$$X_{t} \sim GEV(\mu(t), \sigma, \xi)$$
$$\mu(t)=\beta_{0}+\beta_{1}t$$

```{r warning=FALSE,message=FALSE}
library(ismev)
df<-df_max_rainfall__38002
time<-c(seq(1,length(df$Max_prcp_data_38002),1))
df<-cbind(df,time)
df <- df[!is.infinite(rowSums(df)),]
df1 <- as.matrix(df[,1])
df2 <- as.matrix(df[,2])
A = gev.fit(df1,ydat = df2,mul = 1)
gev.diag(A)
```

#### Significance of stationary and non-stationary GEV distribution model.

* $D= 2[nlls − nllns]=2[-801.2452-(-799.3186)]=-3.8532$

* Chi-squre test value at 5% significance level

```{r warning=FALSE,message=FALSE}
# Upper tail critical value
qchisq(p=0.975, df=1, lower.tail=TRUE)
```

D value is not larger than Chi square value. Hence, non-stationary model does not explain more variability in data than stationary model. 

##### Case 3
$$X_{t} \sim GEV(\mu, \sigma(t), \xi)$$
$$\sigma(t)=exp[\beta_{0}+\beta_{1}t]$$


```{r warning=FALSE,message=FALSE}
library(ismev)
df<-df_max_rainfall__38002
time<-c(seq(1,length(df$Max_prcp_data_38002),1))
df<-cbind(df,time)
df <- df[!is.infinite(rowSums(df)),]
df1 <- as.matrix(df[,1])
df2 <- as.matrix(df[,2])
A = gev.fit(df1,ydat = df2,sigl = 1, siglink = exp)
gev.diag(A)
```

##### Case 4
$$X_{t} \sim GEV(\mu, \sigma(t), \xi)$$
$$\mu(t)=\beta_{0}+\beta_{1}t$$
$$\sigma(t)=exp[\beta_{0}+\beta_{1}t]$$


```{r warning=FALSE,message=FALSE}
library(ismev)
df<-df_max_rainfall__38002
time<-c(seq(1,length(df$Max_prcp_data_38002),1))
df<-cbind(df,time)
df <- df[!is.infinite(rowSums(df)),]
df1 <- as.matrix(df[,1])
df2 <- as.matrix(df[,2])
A = gev.fit(df1,ydat = df2,mul=1, sigl = 1, siglink = exp)
gev.diag(A)
```

#### Gauge Number: 9519

##### Case 1

```{r warning=FALSE,message=FALSE}
df<-Max_prcp_data_9519
df <- df[!is.infinite(rowSums(df)),]
A = gev.fit(df,ydat = NULL)
gev.diag(A)

```

##### Case 2

```{r warning=FALSE,message=FALSE}
df<-df_max_rainfall__9519
time<-c(seq(1,length(df$Max_prcp_data_9519),1))
df<-cbind(df,time)
df <- df[!is.infinite(rowSums(df)),]
df1 <- as.matrix(df[,1])
df2 <- as.matrix(df[,2])
A = gev.fit(df1,ydat = df2,mul = 1)
gev.diag(A)
```

##### Case 3


```{r warning=FALSE,message=FALSE}
library(ismev)
df<-df_max_rainfall__9519
time<-c(seq(1,length(df$Max_prcp_data_9519),1))
df<-cbind(df,time)
df <- df[!is.infinite(rowSums(df)),]
df1 <- as.matrix(df[,1])
df2 <- as.matrix(df[,2])
A = gev.fit(df1,ydat = df2,sigl = 1, siglink = exp)
gev.diag(A)
```

##### Case 4


```{r warning=FALSE,message=FALSE}
df<-df_max_rainfall__9519
time<-c(seq(1,length(df$Max_prcp_data_9519),1))
df<-cbind(df,time)
df <- df[!is.infinite(rowSums(df)),]
df1 <- as.matrix(df[,1])
df2 <- as.matrix(df[,2])
A = gev.fit(df1,ydat = df2,mul=1, sigl = 1, siglink = exp)
gev.diag(A)
```

# **Peaks Over Threshold (POT) Approach - Generalized Pareto distribution (GPD)**

* Threshold value is decided based on 95th quantile

```{r warning=FALSE,message=FALSE}
# Quantile value finding
prcp_rank_data_38002<-sort(prcp_data_38002$prcp)
threshold_38002<-quantile(prcp_rank_data_38002, probs = 0.95)
threshold_38002
```

#### Other ways to select threshold

* A mean residual life plot

* **Select the smallest threshold above which the mean excess plot is roughly linear.**

```{r warning=FALSE,message=FALSE}
data <- prcp_data_38002$prcp
data<- na.omit(data)
mrl.plot(data,umax = 100)
```

* **selecting the threshold is based on GPD parameters which are constant with U>Uo, where Uo is the threshold value.**

```{r warning=FALSE,message=FALSE}
data <- prcp_data_38002$prcp
data<- na.omit(data)
gpd.fitrange(data,0 , 20 , nint=100)

```


Therefore, rainfall values greater than "threshold_38002" is used to select extreme rainfall data.

```{r warning=FALSE,message=FALSE}
Extreme_rainfall_38002<-prcp_data_38002 %>% filter(prcp > threshold_38002)
```

```{r warning=FALSE,message=FALSE}
ggplot(Extreme_rainfall_38002, aes(x=date, y=prcp))+ geom_line(col="#1b9e77")+ggtitle("Gauge number:38002-Extreme rainfall by PoT method")+ theme(plot.title=element_text(hjust=0.5))
```

Generalized Pareto Distribution modeling

##### Case 1
$$X_{t} \sim GPD(\sigma, \xi)$$

```{r warning=FALSE,message=FALSE}
df<-prcp_data_38002
df<-df[,c(2,3)]
df$date<-lubridate::year(df$date)
df <- na.omit(df)
B = gpd.fit(df$prcp,threshold = 8)
gpd.diag(B)

```


##### Case 2
$$X_{t} \sim GPD(\sigma(t), \xi)$$
$$\sigma(t)=exp[\beta_{0}+\beta_{1}t]$$


```{r warning=FALSE,message=FALSE}
df<-prcp_data_38002
df<-df[,c(2,3)]
df$date<-lubridate::year(df$date)
df <- na.omit(df)
df1<- as.matrix(df$date)
B = gpd.fit(df$prcp,threshold = 8,ydat = df1,sigl = 1, siglink = exp)
gpd.diag(B)

```

##### Case 3

$$X_{t} \sim GPD(\sigma(t), \xi (t))$$

$$\sigma(t)=exp[\beta_{0}+\beta_{1}t]$$

$$\xi(t)=\beta_{0}+\beta_{1}t$$


```{r warning=FALSE,message=FALSE}
df<-prcp_data_38002
df<-df[,c(2,3)]
df$date<-lubridate::year(df$date)
df <- na.omit(df)
df1<- as.matrix(df$date)
B = gpd.fit(df$prcp,threshold = 8,ydat = df1,sigl = 1, siglink = exp, shl = 1)
gpd.diag(B)

```

